---
title: Clustering data
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(seafisheries)
library(dplyr)
library(ggplot2)
library(lubridate)
library(here)

dir.create(here("outputs"), showWarnings = FALSE, recursive = TRUE)
stat_list <- list()
```

This document demonstrates clustering of operational longline data to identify
fishing strategies. The input data comes from the cleaned operational data
(see vignette 02).

Summary:

- Selects relevant variables for clustering
- Performs PCA to reduce dimensionality
- Uses gap statistic to determine optimal number of clusters
- Assigns clusters to all data points
- Merges small clusters based on catch threshold

# Load cleaned data

```{r}
EC_clean <- readRDS(here("outputs/EC_data.RDS"))
stat_list$nrow_total <- nrow(EC_clean)
```

```{r}
# Will be in another tutorial soon
EC_clean <- EC_clean %>%
	mutate(
		total_n      = yft_n + alb_n + bet_n,
		yft_fraction = ifelse(total_n > 0, yft_n / total_n, 0),
		alb_fraction = ifelse(total_n > 0, alb_n / total_n, 0),
		bet_fraction = ifelse(total_n > 0, bet_n / total_n, 0)
	) %>%
	rename(mean_hbf = hbf) %>%
	filter(!is.na(mean_hbf))
```

# Define clustering parameters

```{r}
# Clustering scenario: yellowfin, bigeye, albacore + hooks between floats
scenario <- "yba_hbf"

# PCA and kmeans parameters
pca_variance_threshold <- 0.7   # Retain PCs explaining 70% of variance
sample_no              <- 100   # Sample size per replicate for gap statistic
kmeans_set             <- 15    # Number of kmeans replicates
max_k                  <- 20    # Maximum number of clusters to test
iter_max               <- 1e6   # Maximum iterations for kmeans
nstart                 <- 10    # Number of random starts for kmeans
B_var                  <- 25    # Bootstrap replicates for gap statistic

# Cluster merging parameters
catch_threshold <- 3            # Merge clusters with < 3% of total catch
```

# Select and prepare variables

```{r}
pca_select <- selectScenario(scenario)

cat("Variables selected for clustering:\n")
print(pca_select)

select_dat <- EC_clean[, pca_select]
pca_scaled <- pcaScale(select_dat, method = "zscore")
```

# Principal Component Analysis

```{r}
pca_res <- myPCA(pca_scaled,
				 variance_threshold = pca_variance_threshold,
				 print_it = FALSE)

cat("PCA retained", pca_res$no_var, "components explaining >=",
	pca_variance_threshold * 100, "% of variance\n\n")

pca_full <- pca_res$x[, 1:pca_res$no_var]

# Filter to post-1990 data for clustering
indices_90  <- which(year(EC_clean$ymd) >= 1990)
pca_full_90 <- pca_full[indices_90, ]

cat("Using", nrow(pca_full_90), "observations from 1990 onwards for clustering\n")
stat_list$nrow_clustering <- nrow(pca_full_90)
```

# Determine optimal number of clusters

```{r}
set.seed(123)
sample_indices <- lapply(1:kmeans_set, function(x) {
	sample(1:nrow(pca_full_90), size = sample_no, replace = TRUE)
})

cat("\nRunning", kmeans_set, "kmeans replicates on samples of",
	sample_no, "observations...\n")

kmeans_res <- list()
gap_stat   <- list()

for (i in 1:kmeans_set) {
	pca_sub <- pca_full_90[sample_indices[[i]], 1:pca_res$no_var]

	res <- customKmeans(pca_sub,
						max_k      = max_k,
						random_set = B_var,
						iter_max   = iter_max,
						nstart     = nstart,
						d.power    = 2,
						print_it   = FALSE)

	kmeans_res[[i]] <- res$kmeans
	gap_stat[[i]]   <- res$gap_stat
}

# Determine most common K across replicates
cluster_counts <- data.frame(
	cluster_n = sapply(kmeans_res, function(x) length(unique(x$cluster))),
	replicate = 1:kmeans_set
)

freq_table    <- table(cluster_counts$cluster_n)
main_clust_n  <- as.numeric(names(freq_table)[which.max(freq_table)])

cat("\nCluster number distribution across replicates:\n")
print(freq_table)
cat("\nSelected K =", main_clust_n, "clusters\n")
stat_list$selected_k <- main_clust_n
```

## Standardize cluster number across replicates

```{r}
samples_to_rerun <- which(cluster_counts$cluster_n != main_clust_n)

if (length(samples_to_rerun) > 0) {
	cat("\nRerunning", length(samples_to_rerun),
		"replicates with K =", main_clust_n, "\n")

	for (i in samples_to_rerun) {
		pca_sub        <- pca_full_90[sample_indices[[i]], 1:pca_res$no_var]
		kmeans_res[[i]] <- kmeans(pca_sub,
								  centers  = main_clust_n,
								  iter.max = iter_max,
								  nstart   = nstart)
	}
}
```

# Apply clustering to full dataset

```{r}
cat("\nRunning kmeans on full dataset with K =", main_clust_n, "\n")
kmeans_full <- kmeans(pca_full_90,
					  centers  = main_clust_n,
					  iter.max = iter_max,
					  nstart   = nstart)

EC_clean$cluster <- assignClusters(as.data.frame(pca_full),
								   as.data.frame(kmeans_full$centers))

cat("Assigned clusters to", nrow(EC_clean), "observations\n")
```

# Merge small clusters

```{r}
cluster_summary <- summaryClusters(EC_clean)

cat("\nInitial cluster summary:\n")
print(knitr::kable(cluster_summary[, c("cluster", "yft_catch", "yft_frac")],
				   col.names = c("Cluster", "% Total Catch", "% YFT in Cluster"),
				   digits = 1))

small_clusters <- cluster_summary %>%
	filter(yft_catch < catch_threshold)

if (nrow(small_clusters) > 0) {
	cat("\nClusters below", catch_threshold, "% catch threshold:\n")
	print(small_clusters$cluster)

	main_clusters  <- cluster_summary %>%
		filter(yft_catch >= catch_threshold) %>%
		pull(cluster)

	main_centroids <- as.data.frame(kmeans_full$centers)[main_clusters, ]

	EC_clustered <- EC_clean %>%
		mutate(cluster = assignClusters(as.data.frame(pca_full),
										main_centroids)) %>%
		orderClusters()

	stat_list$n_clusters_final <- length(unique(EC_clustered$cluster))

} else {
	cat("\nNo clusters below threshold\n")
	EC_clustered <- EC_clean %>% orderClusters()
	stat_list$n_clusters_final <- main_clust_n
}

cat("\nFinal number of clusters:", stat_list$n_clusters_final, "\n")
```

## Final cluster summary

```{r}
final_summary <- summaryClusters(EC_clustered)

knitr::kable(
	final_summary[, c("cluster", "yft_n", "total_n", "yft_catch", "yft_frac")],
	format     = "simple",
	col.names  = c("Cluster", "YFT Caught", "Total Catch",
				   "% of Total", "% YFT in Cluster"),
	caption    = "Final Cluster Summary",
	digits     = c(0, 0, 0, 1, 1)
)
```

# Save clustered data

```{r}
saveRDS(EC_clustered, here("outputs/EC_data_clustered.RDS"))
cat("\nClustering complete. Data saved to outputs/EC_data_clustered.RDS\n")
```
