---
title: 04 Clustering diagnostic report
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04 Clustering diagnostic report}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(seafisheries)
library(dplyr)
library(ggplot2)
library(lubridate)
library(cluster)
library(here)

# Load clustered data and diagnostics produced by clusters.qmd
EC_clustered       <- readRDS(here("outputs/EC_data_clustered.RDS"))
cluster_diagnostics <- readRDS(here("outputs/cluster_diagnostics.RDS"))

# Unpack diagnostics into the local environment
list2env(cluster_diagnostics, envir = environment())
# Available: pca_res, pca_select, pca_variance_threshold, scenario,
#            gap_stat, kmeans_res, sample_indices, freq_df, stat_list

# Recompute scaled PCA input from the saved clustered data.
# EC_clustered retains all original columns, so pca_select still applies.
select_dat <- EC_clustered[, pca_select]
pca_scaled <- pcaScale(select_dat, method = "zscore")

# Project through the saved pca_res (avoids re-fitting; results are identical
# to the original run as long as EC_clustered hasn't changed).
pca_full    <- predict(pca_res, newdata = pca_scaled)
indices_90  <- which(year(EC_clustered$ymd) >= 1990)
pca_full_90 <- pca_full[indices_90, 1:pca_res$no_var]

kmeans_set <- length(kmeans_res)
max_k      <- nrow(gap_stat[[1]])
```

This report summarises the clustering run produced by vignette 03.
It loads pre-computed objects directly — no clustering is re-run here.

# Run parameters

```{r params_table}
params_df <- data.frame(
	Parameter = c("Scenario", "PCA variance threshold", "Sample size per replicate",
				  "K-means replicates", "Max K tested", "Bootstrap replicates (gap)",
				  "Catch threshold for merging (%)"),
	Value = c(scenario,
			  pca_variance_threshold,
			  stat_list$nrow_clustering / kmeans_set,  # sample_no not stored; derive
			  kmeans_set,
			  max_k,
			  length(gap_stat[[1]]),                   # B_var = nrow of gap mat - max_k
			  NA)                                       # catch_threshold not stored
)
# Note: sample_no and catch_threshold are not in cluster_diagnostics.
# Add them to stat_list in clusters.qmd if you need them here.
knitr::kable(params_df[!is.na(params_df$Value), ], row.names = FALSE,
			 caption = "Clustering parameters")
```

```{r params_counts}
counts_df <- data.frame(
	Item = c("Total observations", "Observations used for clustering (>=1990)",
			 "K selected", "Final number of clusters"),
	Value = c(stat_list$nrow_total,
			  stat_list$nrow_clustering,
			  stat_list$selected_k,
			  stat_list$n_clusters_final)
)
knitr::kable(counts_df, row.names = FALSE, caption = "Data and cluster counts")
```

# PCA

## Loadings

```{r pca_loadings}
# Loadings are available directly from the saved pca_res object.
loadings_df <- as.data.frame(pca_res$rotation[, 1:pca_res$no_var])
knitr::kable(round(loadings_df, 3), caption = "PCA loadings (retained components)")
```

## Variance explained

```{r pca_variance, fig.height=4}
var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2)
cumvar        <- cumsum(var_explained)

var_df <- data.frame(
	PC      = seq_along(var_explained),
	var     = var_explained,
	cum_var = cumvar
)

ggplot(var_df, aes(x = PC)) +
	geom_col(aes(y = var), fill = "steelblue", alpha = 0.7) +
	geom_line(aes(y = cum_var), colour = "firebrick", linewidth = 0.8) +
	geom_point(aes(y = cum_var), colour = "firebrick", size = 2) +
	geom_hline(yintercept = pca_variance_threshold,
			   linetype = "dashed", colour = "grey40") +
	scale_y_continuous(labels = scales::percent_format()) +
	labs(x = "Principal component", y = "Variance explained",
		 title = sprintf("Cumulative variance — %d PCs retained at %.0f%% threshold",
		 				pca_res$no_var, pca_variance_threshold * 100)) +
	theme_bw()
```

## Diagnostic plots

```{r pca_diag_plots}
# Assumption: myPCA() with print_it = TRUE generates a loading biplot and
# a PC scatter plot inline. If your version uses return_print = TRUE instead,
# replace this call with:
#   pca_prints <- myPCA(pca_scaled, variance_threshold = pca_variance_threshold,
#                        print_it = FALSE, return_print = TRUE)
#   print(pca_prints$loading_plot)
#   print(pca_prints$scatter_plot)
myPCA(pca_scaled, variance_threshold = pca_variance_threshold, print_it = TRUE)
```

# Gap statistics

## Cluster count frequency across replicates

```{r gap_freq_table}
knitr::kable(
	freq_df %>% arrange(desc(freq)),
	col.names = c("N clusters", "Frequency"),
	caption   = sprintf("K selected = %d (most frequent across %d replicates)",
						stat_list$selected_k, kmeans_set)
)
```

## Gap statistic plots

One plot per replicate. The vertical line marks the K selected by the
Tibshirani 2001 SE rule.

```{r gap_plots, fig.height=3, fig.width=5, results='hide'}
for (i in seq_along(gap_stat)) {
	nc <- cluster::maxSE(
		f        = gap_stat[[i]][, "gap"],
		SE.f     = gap_stat[[i]][, "SE.sim"],
		method   = "Tibs2001SEmax",
		SE.factor = 1
	)
	plot_gap <- data.frame(
		k   = factor(seq_len(max_k)),
		gap = gap_stat[[i]][, "gap"],
		sd  = gap_stat[[i]][, "SE.sim"]
	)
	p <- ggplot(plot_gap, aes(x = k, y = gap)) +
		geom_point() +
		geom_errorbar(aes(ymin = gap - sd, ymax = gap + sd), width = 0.3) +
		geom_vline(xintercept = nc, colour = "firebrick", linetype = "dashed") +
		labs(title = sprintf("Replicate %d  —  K = %d", i, nc),
			 x = "K", y = "Gap statistic") +
		theme_bw()
	print(p)
}
```

# Clustering performance

```{r perf_setup}
# Extract final cluster labels for the >=1990 subset
cluster_90 <- EC_clustered$cluster[indices_90]
nc         <- length(unique(cluster_90))
```

## Silhouette indices

Silhouette computed on each replicate's sample (same `sample_indices` used
during clustering), averaged per cluster.

```{r silhouette_table}
if (nc > 1) {
	sil_avg <- do.call(rbind, lapply(seq_len(kmeans_set), function(i) {
		pca_sub  <- pca_full_90[sample_indices[[i]], ]
		clusters <- as.integer(cluster_90[sample_indices[[i]]])
		sil      <- cluster::silhouette(x    = clusters,
										dist = dist(pca_sub, method = "euclidean"))
		as.data.frame(sil) %>%
			mutate(sample = i)
	})) %>%
		group_by(cluster) %>%
		summarise(sil_avg = mean(sil_width),
				  sil_sd  = sd(sil_width),
				  .groups = "drop")

	knitr::kable(sil_avg,
				 col.names = c("Cluster", "Mean silhouette", "SD"),
				 digits    = 3,
				 caption   = "Average silhouette width per cluster across replicates")
} else {
	cat("Single cluster — silhouette not applicable.\n")
}
```

```{r silhouette_plot, fig.height=4}
if (nc > 1) {
	ggplot(sil_avg, aes(x = factor(cluster), y = sil_avg)) +
		geom_point(size = 3) +
		geom_errorbar(aes(ymin = sil_avg - sil_sd,
						  ymax = sil_avg + sil_sd), width = 0.2) +
		geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
		labs(x = "Cluster", y = "Mean silhouette width") +
		theme_bw()
}
```

## Dunn index

Averaged across replicates. Requires the `clValid` package
(`install.packages("clValid")` — add to `Suggests` in DESCRIPTION).

```{r dunn}
if (nc > 1) {
	if (!requireNamespace("clValid", quietly = TRUE)) {
		cat("clValid not installed — Dunn index skipped.\n")
	} else {
		dunn_indices <- vapply(seq_len(kmeans_set), function(i) {
			pca_sub  <- pca_full_90[sample_indices[[i]], ]
			clusters <- as.integer(cluster_90[sample_indices[[i]]])
			clValid::dunn(distance = dist(pca_sub, method = "euclidean"),
						  clusters = clusters)
		}, numeric(1))

		cat(sprintf("Dunn index: mean = %.4f  (sd = %.4f, n = %d replicates)\n",
					mean(dunn_indices), sd(dunn_indices), kmeans_set))
	}
} else {
	cat("Single cluster — Dunn index not applicable.\n")
}
```

# Cluster assignment before and after merging

`cluster_init` holds the raw k-means labels before small clusters were merged
into the nearest large one. `cluster` holds the final, ordered labels.

## Variable distributions — before merging

```{r before_merge_summary}
EC_before <- EC_clustered %>% mutate(cluster = cluster_init)
table_before <- summaryClusters(EC_before)

split_at <- floor(ncol(table_before) / 2) + 1
knitr::kable(table_before[, 1:split_at],
			 digits  = 1,
			 caption = "Before merging (part 1)")
knitr::kable(table_before[, c(1, (split_at + 1):ncol(table_before))],
			 digits  = 1,
			 caption = "Before merging (part 2)")
```

## Variable distributions — after merging

```{r after_merge_summary}
table_after <- summaryClusters(EC_clustered)

knitr::kable(table_after[, 1:split_at],
			 digits  = 1,
			 caption = "After merging (part 1)")
knitr::kable(table_after[, c(1, (split_at + 1):ncol(table_after))],
			 digits  = 1,
			 caption = "After merging (part 2)")
```

## Species catch proportions — before merging

```{r species_before}
EC_before %>%
	group_by(cluster) %>%
	summarise(cluster_size = n(),
			  yft_n        = sum(yft_n),
			  bet_n        = sum(bet_n),
			  alb_n        = sum(alb_n),
			  total_n      = sum(total_n),
			  .groups      = "drop") %>%
	mutate(prop       = cluster_size / sum(cluster_size),
		   prop_yft   = yft_n   / sum(yft_n),
		   prop_bet   = bet_n   / sum(bet_n),
		   prop_alb   = alb_n   / sum(alb_n),
		   prop_total = total_n / sum(total_n)) %>%
	mutate(across(starts_with("prop"), ~round(.x, 2))) %>%
	select(cluster, cluster_size, prop,
		   yft_n, prop_yft, bet_n, prop_bet, alb_n, prop_alb,
		   total_n, prop_total) %>%
	knitr::kable(caption = "Species proportions before merging")
```

## Species catch proportions — after merging

```{r species_after}
EC_clustered %>%
	group_by(cluster) %>%
	summarise(cluster_size = n(),
			  yft_n        = sum(yft_n),
			  bet_n        = sum(bet_n),
			  alb_n        = sum(alb_n),
			  total_n      = sum(total_n),
			  .groups      = "drop") %>%
	mutate(prop       = cluster_size / sum(cluster_size),
		   prop_yft   = yft_n   / sum(yft_n),
		   prop_bet   = bet_n   / sum(bet_n),
		   prop_alb   = alb_n   / sum(alb_n),
		   prop_total = total_n / sum(total_n)) %>%
	mutate(across(starts_with("prop"), ~round(.x, 2))) %>%
	select(cluster, cluster_size, prop,
		   yft_n, prop_yft, bet_n, prop_bet, alb_n, prop_alb,
		   total_n, prop_total) %>%
	knitr::kable(caption = "Species proportions after merging")
```
