---
title: "03 Processing Length Frequency Data"
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03 Processing Length Frequency Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, warning=FALSE, message=F, echo=FALSE}
library(seafisheries)
library(dplyr)
stat_list <- list()
library(here)
dir.create(here("outputs"), showWarnings = FALSE, recursive = TRUE)
```

This document processes raised 5x5 length-frequency data. Unlike catch/effort
data, gear type doesn't change the cleaning logic here, so there's no `gear`
argument anywhere in this pipeline.

## Load data and rename columns

Renaming is hardcoded per source, same as the operational data pipeline. 
This is metadata about *a specific* dataset, not something a function
can infer.

```{r}
LF_dat <- readRDS(paste0(here(),"/inst/extdata/mockData/length_frequency_data.RDS"))

LF_dat <- LF_dat %>% 
	rename(id = ID,
		   yr = YR,
		   qtr = QTR,
		   mm = MON,
		   gr = GR,
		   flag = FLAG_ID,
		   fleet = FLEET_ID,
		   latitude = LAT_SHORT,
		   longitude = LON_SHORT,
		   len = LEN,
		   count = FREQ,
		   school_LF = SCHTYPE_ID) %>% 
	mutate(len = as.numeric(len))

# ASTRAT and LSTRAT are kept under their original names
```

## Reference table: ASTRAT grid resolution

`ASTRAT` encodes which grid resolution a record was aggregated to, and that
resolution determines the offset added to raw coordinates to get a cell
center. This table is specific to this data source and lives here, not in
the package. `compute_grid_center()` (below) is generic and takes this as
a plain argument.

```{r}
astrat_lookup <- data.frame(
	ASTRAT     = c("5",  "0", "F",  "T", "A", "1"),
	lon_offset = c(2.5,  5,   5,    10, 5, 0.5),
	lat_offset = c(2.5,  5,   2.5,  5, 2, 0.5)
)

valid_astrat <- astrat_lookup$ASTRAT

knitr::kable(astrat_lookup, caption = "ASTRAT resolution reference")
```

## Convert coordinates from degree-minutes to decimal degrees

Same package function used for operational data; gear-agnostic, works on
any lat/lon columns regardless of source.

```{r}
res <- convert_coordinates_df(LF_dat, lat_col = "latitude", lon_col = "longitude", report = TRUE)
LF_dat <- res$data
print(res$summary)
```


## Fill in missing month values

`fill_missing_month()` derives `mm` from `qtr` (Q1->Jan, Q2->Apr, Q3->Jul,
Q4->Oct) wherever `mm` is missing or invalid (here, `mm == 0`). If the
column didn't exist at all it would be created entirely from `qtr`.

```{r}
LF_dat <- fill_missing_month(LF_dat, mm_col = "mm", qtr_col = "qtr")
```



## Remove invalid entries

`remove_invalid()` drops rows with NA, NaN, Inf, or NULL in any of the
columns considered essential here: coordinates and length. `ASTRAT`
validity is handled separately below, not here.

```{r}
required_cols <- c("latitude", "longitude", "len")

LF_dat <- remove_invalid(LF_dat, required_cols)
```

## Remove invalid ASTRAT values

`filter_valid_categories()` drops rows whose `ASTRAT` isn't one of the
resolutions in the reference table above. This also removes NA `ASTRAT`
values, since `NA %in% valid_astrat` is FALSE.

```{r}
LF_dat <- filter_valid_categories(LF_dat, col = "ASTRAT", valid_values = valid_astrat)
```

## Compute grid-cell centers

`compute_grid_center()` joins the ASTRAT reference table and adds the
matching lon/lat offset, producing `lonCent`/`latCent`. Every `ASTRAT` value
remaining at this point is already valid (previous step), so there should be
no unmatched rows.

```{r}
LF_dat <- compute_grid_center(LF_dat, strat_col = "ASTRAT", lookup = astrat_lookup,
							   lon_col = "longitude", lat_col = "latitude")
```

## Apply the bounding box

`filter_bounding_box()` is the LF equivalent of `apply_pacific_mask()` used
for operational data but without the `pmask_lookup` semi-join, since that
lookup is built for the 1x1 grid and won't match ASTRAT's coarser cells.

```{r}
LF_dat <- filter_bounding_box(LF_dat, lat_col = "latCent", lon_col = "lonCent",
							   lat_range = lat_range, lon_range = lon_range)
```

## Remove sparse samples

`remove_sparse_lf_samples()` drops samples (grouped by the columns below)
with fewer than 3 distinct length bins, or where every bin has the same
count. Both cases carry no usable length-distribution information.

```{r}
group_cols <- c("yr", "qtr", "gr", "flag", "latCent", "lonCent", "ASTRAT", "school_LF")

LF_dat <- remove_sparse_lf_samples(LF_dat, group_cols = group_cols, 
									len_col = "len", freq_col = "count", min_bins = 3)
```


## Convert and clean school-association codes (for Purse Seine only)

Length-frequency data encodes school association as letter codes, while
catch/effort data uses numeric codes 1-7 (see `treat_school()`). The two
schemes don't line up one-to-one. LF data doesn't distinguish the
sub-categories the numeric scheme allows for some codes, so two of the six
mappings below collapse multiple numeric values into one:

```{r, echo = FALSE}
school_lookup_table <- data.frame(
	school_type_id = c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8),
	description = c(
		"Unknown",
		"No school type (only searching activity)",
		"Unassociated/free school",
		"Feeding on baitfish",
		"Drifting log, debris or dead animal",
		"Drifting raft, FAD or payao",
		"Anchored raft, FAD or payao",
		"Live whale",
		"Live whale shark",
		"Other"
	),
	"LF code" = c("", "", "U", "U", "L", "F", "F", "M", "A", "O"),
	check.names = FALSE
)
knitr::kable(school_lookup_table, caption = "School-association code conversion (catch/effort -> LF)")
```

`F` and `U` each cover two numeric codes in the catch/effort scheme; since
LF data can't distinguish between them, `F` always maps to `4` and `U`
always maps to `1`, by convention rather than by any real distinction in
the data. `0` (unknown) and anything not in this table map straight to
`10`, matching how an out-of-range value would be handled if it showed up
directly in catch/effort data.

`convert_school_letters()` creates a new numeric `school` column from the
raw `school_LF` letter codes, keeping `school_LF` untouched so the original
code is still there to check against later. `treat_school()` then runs on
`school` and catches anything `convert_school_letters()` couldn't map
(left as NA) plus any numeric value that's still outside 1-7, using the
same logic already applied to catch/effort data.

```{r}
LF_dat <- convert_school_letters(LF_dat, school_col = "school_LF", output_col = "school")
LF_dat <- treat_school(LF_dat, school_col = "school")
```


## All of the above, in one call

```{r}
LF_dat_clean <- process_lf_data(
	df            = LF_dat,
	required_cols = required_cols,
	mm_col        = "mm",
	qtr_col       = "qtr",
	strat_col     = "ASTRAT",
	valid_strat   = valid_astrat,
	strat_lookup  = astrat_lookup,
	lon_col       = "longitude",
	lat_col       = "latitude",
	lat_range     = lat_range,
	lon_range     = lon_range,
	group_cols    = group_cols,
	len_col       = "len",
	freq_col      = "count",
	min_bins      = 3
)
```

Note this chunk reruns the whole pipeline on `LF_dat`, which the individual
sections above already cleaned. This is a side-by-side reference, not a
script meant to run start to finish; reload raw data first if you want to
actually execute the wrapper version.

```{r}
saveRDS(LF_dat_clean, here("outputs", "LF_clean.RDS"))
```

## Disaggregate to 1cm bins

Separate from cleaning. `bin_to_1cm()` expands each row's `LSTRAT`-sized
bin into individual 1cm bins, splitting `count` proportionally. Run this
after cleaning, on the cleaned data, not as part of `process_lf_data()`.

```{r}
LF_uniform_bins <- bin_to_1cm(LF_dat_clean)
saveRDS(LF_uniform_bins, here("outputs", "LF_5x5_bin1.RDS"))
```
