---
title: To create mock up data to use in the other documents
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{To create mock up data to use in the other documents}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = F, warning = F, message = F, warning = FALSE}
knitr::opts_chunk$set(eval = T, echo = F, message = F, warning = F)

```

# Operational data

```{r}
# Helper to convert lat/lon back to DDDMM format
decimal_to_dddmm <- function(decimal, is_lon = FALSE) {
  if (is.na(decimal)) return(NA)
  
  if (is_lon && decimal > 180) {
    # Was originally West, convert back
    decimal <- 360 - decimal
    direction <- "W"
  } else if (is_lon) {
    direction <- "E"
  } else {
    direction <- ifelse(decimal < 0, "S", "N")
  }
  
  degrees <- floor(abs(decimal))
  minutes <- (abs(decimal) - degrees) * 60
  
  # Format: DDDMM.MMM or DDMM.MMM
  if (is_lon) {
    dddmm <- sprintf("%03d%05.2f", degrees, minutes)
  } else {
    dddmm <- sprintf("%02d%05.2f", degrees, minutes)
  }
  
  paste0(dddmm, direction)
}

# Generate synthetic captain logsheet data
generate_mock_captain_LL <- function(n_total = 500, stats = NULL) {
  
  # If no stats provided, use reasonable defaults based on lon_range/lat_range
  if (is.null(stats)) {
    stats <- list(
      date_min = as.Date("1960-01-01"),
      date_max = as.Date("2023-12-31"),
      lat_min = -54.5, lat_max = 64.5,
      lon_min = 89.5, lon_max = 288.5,
      effort_mean = 20, effort_sd = 10,
      hbf_mean = 15, hbf_sd = 5,
      catch_mean = 50, catch_sd = 30,
      weight_mean = 2, weight_sd = 1,
      flags = c("JP", "TW", "CN", "US", "AU")
    )
  }
  
  vessel_names <- c("Setosa_I", "Setosa_II", "Versicolor_I", "Versicolor_II", 
                    "Virginica_I", "Virginica_II", "Setosa_III", "Versicolor_III")
  
  # Category sizes
  n_good <- 1335
  n_na <- 30
  n_hbf_outliers <- 20
  n_dummy <- 15
  n_zero <- 20
  n_effort_outliers <- 15
  n_bad_names <- 10
  n_duplicates <- 20
  n_outside_mask <- 20
  n_coord_errors <- 15
  
  # Generate base "good" data
  good_data <- data.frame(
    flag = sample(stats$flags, n_good, replace = TRUE),
    vesselname = sample(vessel_names, n_good, replace = TRUE),
    set_date = sample(seq(stats$date_min, stats$date_max, by = "day"), n_good, replace = TRUE),
    latitude = runif(n_good, stats$lat_min, stats$lat_max),
    longitude = runif(n_good, stats$lon_min, stats$lon_max),
    E = abs(rnorm(n_good, stats$effort_mean, stats$effort_sd)),
    hbf = abs(rnorm(n_good, stats$hbf_mean, stats$hbf_sd)),
    yft_n = rpois(n_good, stats$catch_mean),
    yft_w = abs(rnorm(n_good, stats$weight_mean, stats$weight_sd)),
    skj_n = rpois(n_good, stats$catch_mean * 0.3),
    skj_w = abs(rnorm(n_good, stats$weight_mean * 0.3, stats$weight_sd)),
    bet_n = rpois(n_good, stats$catch_mean * 0.5),
    bet_w = abs(rnorm(n_good, stats$weight_mean * 0.5, stats$weight_sd)),
    alb_n = rpois(n_good, stats$catch_mean * 0.2),
    alb_w = abs(rnorm(n_good, stats$weight_mean * 0.2, stats$weight_sd)),
    oth_n = rpois(n_good, stats$catch_mean * 0.1),
    oth_w = abs(rnorm(n_good, stats$weight_mean * 0.1, stats$weight_sd))
  )
  
  # Convert lat/lon to DDDMM format
  good_data$latitude_dddmm <- sapply(good_data$latitude, decimal_to_dddmm, is_lon = FALSE)
  good_data$longitude_dddmm <- sapply(good_data$longitude, decimal_to_dddmm, is_lon = TRUE)
  
  # NA problems
  na_data <- good_data[1:n_na, ]
  na_fields <- c("set_date", "latitude", "longitude", "E", "yft_n", "skj_n", "bet_n", "alb_n", "oth_n")
  for (i in 1:n_na) {
    na_data[i, sample(na_fields, 1)] <- NA
  }
  
  # HBF outliers - way beyond 3 SD
  hbf_outlier_data <- good_data[1:n_hbf_outliers, ]
  hbf_outlier_data$hbf <- stats$hbf_mean + sample(c(-1, 1), n_hbf_outliers, replace = TRUE) * 
    4 * stats$hbf_sd
  
  # Dummy fisheries
  dummy_data <- good_data[1:n_dummy, ]
  dummy_data$E <- 0
  dummy_data$yft_n <- rpois(n_dummy, stats$catch_mean)
  
  # Zero fisheries
  zero_data <- good_data[1:n_zero, ]
  zero_data$E <- 0
  zero_data$yft_n <- 0
  zero_data$skj_n <- 0
  zero_data$bet_n <- 0
  zero_data$alb_n <- 0
  zero_data$oth_n <- 0
  
  # Effort outliers
  effort_outlier_data <- good_data[1:n_effort_outliers, ]
  effort_outlier_data$E <- stats$effort_mean + sample(c(-1, 1), n_effort_outliers, replace = TRUE) * 
    4 * stats$effort_sd
  
  # Bad vessel names
  name_data <- good_data[1:n_bad_names, ]
  name_data$vesselname <- paste0(name_data$vesselname, "  ")
  
  # Exact duplicates
  dup_data <- good_data[1:(n_duplicates/2), ]
  dup_data <- rbind(dup_data, dup_data)
  
  # Outside mask - obviously wrong coordinates
  outside_data <- good_data[1:n_outside_mask, ]
  # lon < 89.5: Indian Ocean
  outside_data$longitude[1:7] <- runif(7, 40, 85)
  # lon > 288.5: Caribbean/Atlantic
  outside_data$longitude[8:14] <- runif(7, 295, 350)
  # lat < -54.5: Antarctica
  outside_data$latitude[15:17] <- runif(3, -70, -60)
  # lat > 64.5: Arctic
  outside_data$latitude[18:20] <- runif(3, 70, 80)
  
  # Reconvert outside coords to DDDMM
  outside_data$latitude_dddmm <- sapply(outside_data$latitude, decimal_to_dddmm, is_lon = FALSE)
  outside_data$longitude_dddmm <- sapply(outside_data$longitude, decimal_to_dddmm, is_lon = TRUE)
  
  # Bad coordinate formats - comprehensive error examples
n_coord_errors <- 15

# Base data for coordinate errors
coord_error_data <- good_data[1:n_coord_errors, ]

# Assign different coordinate errors (will be applied to latitude_dddmm/longitude_dddmm)
coord_error_data$latitude_dddmm <- c(
  "N",                    # cardinal_only
  "4530X",                # unknown_cardinal  
  "45.30.5N",             # multiple_decimals
  "45A30N",               # invalid_format
  "5N",                   # insufficient_digits
  "4565N",                # minutes_out_of_range (65 minutes)
  "4530E",                # wrong_cardinal_for_type (E on latitude)
  "9530N",                # latitude_out_of_range (95 degrees)
  "",                     # empty string (caught as empty_or_null)
  "12.34.56S",            # multiple_decimals
  "3070N",                # minutes_out_of_range (70 minutes)
  "4530W",                # wrong_cardinal_for_type (W on latitude)
  "2X45N",                # invalid_format
  "S",                    # cardinal_only
  "1234Y"                 # unknown_cardinal
)

coord_error_data$longitude_dddmm <- c(
  "E",                    # cardinal_only
  "12025Z",               # unknown_cardinal
  "120.25.5E",            # multiple_decimals
  "120B25W",              # invalid_format
  "25E",                  # insufficient_digits
  "12065W",               # minutes_out_of_range (65 minutes)
  "12025N",               # wrong_cardinal_for_type (N on longitude)
  "36525E",               # longitude_out_of_range (365 degrees)
  "",                     # empty string
  "150.30.25W",           # multiple_decimals
  "17580E",               # minutes_out_of_range (80 minutes)
  "14530S",               # wrong_cardinal_for_type (S on longitude)
  "1X5025E",              # invalid_format
  "W",                    # cardinal_only
  "9999Q"                 # unknown_cardinal
)

  # Combine all
  all_data <- rbind(
    good_data, na_data, hbf_outlier_data, dummy_data, zero_data,
    effort_outlier_data, name_data, dup_data, outside_data, coord_error_data
  )
  
  # Add estimated values
  all_data$yft_n_est <- ifelse(is.na(all_data$yft_n) | all_data$yft_n == 0, 
                                 rpois(nrow(all_data), 5), NA)
  all_data$skj_n_est <- ifelse(is.na(all_data$skj_n) | all_data$skj_n == 0, 
                                 rpois(nrow(all_data), 3), NA)
  all_data$bet_n_est <- ifelse(is.na(all_data$bet_n) | all_data$bet_n == 0, 
                                 rpois(nrow(all_data), 4), NA)
  all_data$alb_n_est <- ifelse(is.na(all_data$alb_n) | all_data$alb_n == 0, 
                                 rpois(nrow(all_data), 2), NA)
  all_data$oth_n_est <- ifelse(is.na(all_data$oth_n) | all_data$oth_n == 0, 
                                 rpois(nrow(all_data), 1), NA)
  
  # Shuffle
  all_data <- all_data[sample(nrow(all_data)), ]
  
  # Format dates as "1960-10-23 00:00:00.000"
  formatted_dates <- format(all_data$set_date, "%Y-%m-%d 00:00:00.000")
  
  # Create final output
  output_data <- data.frame(
    flag = all_data$flag,
    vesselname = all_data$vesselname,
    latitude = all_data$latitude_dddmm,
    longitude = all_data$longitude_dddmm,
    set_date = formatted_dates,
    hooks_nb = all_data$E * 100,
    hooks_between_floats = all_data$hbf,
    yft_sp_weight_kg = all_data$yft_w * 1000,
    yft_sp_nb = all_data$yft_n,
    yft_sp_nb_est = all_data$yft_n_est,
    skj_sp_weight_kg = all_data$skj_w * 1000,
    skj_sp_nb = all_data$skj_n,
    skj_sp_nb_est = all_data$skj_n_est,
    bet_sp_weight_kg = all_data$bet_w * 1000,
    bet_sp_nb = all_data$bet_n,
    bet_sp_nb_est = all_data$bet_n_est,
    alb_sp_weight_kg = all_data$alb_w * 1000,
    alb_sp_nb = all_data$alb_n,
    alb_sp_nb_est = all_data$alb_n_est,
    oth_sp_weight_kg = all_data$oth_w * 1000,
    oth_sp_nb = all_data$oth_n,
    oth_sp_nb_est = all_data$oth_n_est
  )
  
  return(output_data)
}
```

```{r}
test_data <- generate_mock_captain_LL()
saveRDS(object = test_data, file = "mockData/operational_data.RDS")
```

# aggregated length frequency data

# Length frequency data

Dropped the duplicate `decimal_to_dddmm()` definition here since it's
already defined above in the operational data section -- reused as-is.

Structural difference from the operational data generator above: LF
cleaning operates on *samples* (groups of rows sharing
yr/qtr/gr/flag/ASTRAT/school/coordinates), not independent rows --
`remove_sparse_lf_samples()` needs >=3 distinct length bins per group with
non-identical counts to survive. So generation here is sample-based: each
call to `make_lf_sample()` returns several rows sharing one grouping key,
not one row.

```{r}
# One LF sample = several length-bin rows sharing one grouping key.
# outside_box pushes coordinates far enough outside lat_range/lon_range that
# they stay outside even after compute_grid_center()'s largest offset (10
# degrees, for ASTRAT "T") -- margins below are set with that in mind.
make_lf_sample <- function(sample_id, n_bins, astrat, same_count = FALSE,
							mm_zero = FALSE, outside_box = FALSE, stats) {

	yr     <- sample(stats$yr_min:stats$yr_max, 1)
	qtr    <- sample(1:4, 1)
	mm     <- if (mm_zero) 0 else c(1, 4, 7, 10)[qtr]
	gr     <- sample(stats$gears, 1)
	flag   <- sample(stats$flags, 1)
	fleet  <- sample(stats$fleets, 1)
	school <- sample(stats$schools, 1)
	lstrat <- sample(c(1, 1, 1, 2, 5), 1)  # mostly 1, sometimes >1 to exercise bin_to_1cm

	if (outside_box) {
		side <- sample(c("lon_low", "lon_high", "lat_low", "lat_high"), 1)
		lat <- runif(1, stats$lat_min, stats$lat_max)
		lon <- runif(1, stats$lon_min, stats$lon_max)
		if (side == "lon_low")  lon <- runif(1, 40, 75)
		if (side == "lon_high") lon <- runif(1, 300, 350)
		if (side == "lat_low")  lat <- runif(1, -75, -66)
		if (side == "lat_high") lat <- runif(1, 76, 85)
	} else {
		lat <- runif(1, stats$lat_min, stats$lat_max)
		lon <- runif(1, stats$lon_min, stats$lon_max)
	}

	len_vals <- sort(sample(seq(stats$len_min, stats$len_max, by = 2), n_bins))
	if (same_count) {
		counts <- rep(round(abs(rnorm(1, stats$count_mean, stats$count_sd))) + 1, n_bins)
	} else {
		counts <- round(abs(rnorm(n_bins, stats$count_mean, stats$count_sd))) + 1
	}

	data.frame(
		ID         = paste0(sample_id, "_", seq_len(n_bins)),
		YR         = yr,
		QTR        = qtr,
		MON        = mm,
		GR         = gr,
		FLAG_ID    = flag,
		FLEET_ID   = fleet,
		LAT_SHORT  = decimal_to_dddmm(lat, is_lon = FALSE),
		LON_SHORT  = decimal_to_dddmm(lon, is_lon = TRUE),
		LEN        = len_vals,
		LSTRAT     = lstrat,
		FREQ       = counts,
		ASTRAT     = astrat,
		SCHTYPE_ID = school,
		stringsAsFactors = FALSE
	)
}
```

```{r}
generate_mock_LF <- function(stats = NULL) {

	if (is.null(stats)) {
		stats <- list(
			yr_min = 1990, yr_max = 2023,
			lat_min = -54.5, lat_max = 64.5,
			lon_min = 89.5, lon_max = 288.5,
			len_min = 20, len_max = 180,
			count_mean = 15, count_sd = 8,
			flags = c("JP", "TW", "CN", "US", "AU"),
			fleets = c(1, 2, 3),
			gears = c("L", "S", "P"),
			schools = c("F", "L", "U", "M", "A", "0", "N")
		)
	}

	valid_astrat <- c("5", "0", "F", "T")

	# -- category sizes, in samples/groups, not rows --
	n_good        <- 300  # >=3 bins, varying counts, valid everything
	n_bad_astrat  <- 15   # ASTRAT outside {5,0,F,T} -> filter_valid_categories
	n_outside_box <- 20   # coords outside lat_range/lon_range -> filter_bounding_box
	n_mm_zero     <- 25   # MON = 0 -> fill_missing_month
	n_sparse      <- 15   # 1-2 length bins -> remove_sparse_lf_samples
	n_flat_count  <- 15   # >=3 bins, identical count throughout -> remove_sparse_lf_samples

	good_samples <- lapply(seq_len(n_good), function(i) {
		make_lf_sample(paste0("good", i), n_bins = sample(3:6, 1),
						astrat = sample(valid_astrat, 1), stats = stats)
	})

	bad_astrat_samples <- lapply(seq_len(n_bad_astrat), function(i) {
		make_lf_sample(paste0("badastrat", i), n_bins = sample(3:6, 1),
						astrat = sample(c("9", "X", "2", "Z"), 1), stats = stats)
	})

	outside_box_samples <- lapply(seq_len(n_outside_box), function(i) {
		make_lf_sample(paste0("outside", i), n_bins = sample(3:6, 1),
						astrat = sample(valid_astrat, 1), outside_box = TRUE, stats = stats)
	})

	mm_zero_samples <- lapply(seq_len(n_mm_zero), function(i) {
		make_lf_sample(paste0("mmzero", i), n_bins = sample(3:6, 1),
						astrat = sample(valid_astrat, 1), mm_zero = TRUE, stats = stats)
	})

	sparse_samples <- lapply(seq_len(n_sparse), function(i) {
		make_lf_sample(paste0("sparse", i), n_bins = sample(1:2, 1),
						astrat = sample(valid_astrat, 1), stats = stats)
	})

	flat_count_samples <- lapply(seq_len(n_flat_count), function(i) {
		make_lf_sample(paste0("flat", i), n_bins = sample(3:6, 1),
						astrat = sample(valid_astrat, 1), same_count = TRUE, stats = stats)
	})

	all_data <- do.call(rbind, c(good_samples, bad_astrat_samples, outside_box_samples,
								  mm_zero_samples, sparse_samples, flat_count_samples))

	# -- NA injection: knock out len or a coordinate on a handful of rows.
	#    Applied across the whole combined set, so it can land inside a
	#    sparse/flat sample too and compound with that category rather than
	#    staying isolated -- fine for exercising every function, but don't
	#    expect exact per-category removal counts to match the n_* values
	#    above one-for-one.
	n_na <- 20
	na_rows <- sample(nrow(all_data), n_na)
	na_fields <- c("LEN", "LAT_SHORT", "LON_SHORT")
	for (i in na_rows) {
		all_data[i, sample(na_fields, 1)] <- NA
	}

	# -- coordinate format errors: same 15 malformed DDDMM strings as the
	#    operational data generator's coord_error_data above, since
	#    convert_coordinate()'s failure categories don't depend on which
	#    pipeline is calling it. One single-bin sample per string.
	n_coord_errors <- 15
	coord_error_rows <- do.call(rbind, lapply(seq_len(n_coord_errors), function(i) {
		make_lf_sample(paste0("coorderr", i), n_bins = 1,
						astrat = sample(valid_astrat, 1), stats = stats)
	}))

	coord_error_rows$LAT_SHORT <- c(
		"N",                    # cardinal_only
		"4530X",                # unknown_cardinal
		"45.30.5N",             # multiple_decimals
		"45A30N",                # invalid_format
		"5N",                   # insufficient_digits
		"4565N",                # minutes_out_of_range (65 minutes)
		"4530E",                # wrong_cardinal_for_type (E on latitude)
		"9530N",                # latitude_out_of_range (95 degrees)
		"",                     # empty string (empty_or_null)
		"12.34.56S",            # multiple_decimals
		"3070N",                # minutes_out_of_range (70 minutes)
		"4530W",                # wrong_cardinal_for_type (W on latitude)
		"2X45N",                # invalid_format
		"S",                    # cardinal_only
		"1234Y"                 # unknown_cardinal
	)
	coord_error_rows$LON_SHORT <- c(
		"E",                    # cardinal_only
		"12025Z",               # unknown_cardinal
		"120.25.5E",            # multiple_decimals
		"120B25W",              # invalid_format
		"25E",                  # insufficient_digits
		"12065W",               # minutes_out_of_range (65 minutes)
		"12025N",               # wrong_cardinal_for_type (N on longitude)
		"36525E",               # longitude_out_of_range (365 degrees)
		"",                     # empty string
		"150.30.25W",           # multiple_decimals
		"17580E",               # minutes_out_of_range (80 minutes)
		"14530S",               # wrong_cardinal_for_type (S on longitude)
		"1X5025E",              # invalid_format
		"W",                    # cardinal_only
		"9999Q"                 # unknown_cardinal
	)

	all_data <- rbind(all_data, coord_error_rows)

	# Shuffle
	all_data <- all_data[sample(nrow(all_data)), ]
	rownames(all_data) <- NULL

	all_data
}

decimal_to_dddmm <- function(decimal, is_lon = FALSE) {
	if (is.na(decimal)) return(NA)
	if (is_lon && decimal > 180) {
		decimal <- 360 - decimal
		direction <- "W"
	} else if (is_lon) {
		direction <- "E"
	} else {
		direction <- ifelse(decimal < 0, "S", "N")
	}
	degrees <- floor(abs(decimal))
	minutes <- (abs(decimal) - degrees) * 60
	if (is_lon) {
		dddmm <- sprintf("%03d%05.2f", degrees, minutes)
	} else {
		dddmm <- sprintf("%02d%05.2f", degrees, minutes)
	}
	paste0(dddmm, direction)
}
```

```{r}
mock_lf <- generate_mock_LF()
saveRDS(object = mock_lf, file = "inst/extdata/mockData/length_frequency_data.RDS")
```
