---
title: 02 Processing operational data
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02 Processing Operational Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, warning=FALSE, message=F, echo=FALSE}
library(seafisheries)
library(dplyr)
library(stringr)
stat_list <- list()
pmask_lookup <- readRDS(system.file("extdata/outputs/pmask_lookup.RDS",
                                     package = "seafisheries"))
library(here)
dir.create(here("outputs"), showWarnings = FALSE, recursive = TRUE)

```
This document shows the process of formatting operational data to something usable with models. Using as example the yellowfin captain's and observer's logbooks

Summary:

- converts date/long/lat to usable format

- removes NA in the 3 dimensions and effort

- aggregates to 1x1

- crops to pacific region

- splits data between positive catch and effort / 0 catch and 0 effort / positive catch and 0 effort

# Simple example

This uses a fake data set of longline data. This especially contains hooks between floats so skip the hbf outlier section if using other type of data

## Load data and rename columns

```{r}
EC_raw <- readRDS(system.file("extdata/mockData/operational_data.RDS", 
                               package = "seafisheries"))
# EC_raw <- read.csv("/home/romainf/Documents/FisheriesDatabases/operational_captain_longline_logsheet_catch_effort_data_yft_bet_skj_oth.csv")
stat_list$nrow_init <- nrow(EC_raw)

# Rename columns to standard nomenclature
EC_raw <- EC_raw |> 
	rename(yft_w = yft_sp_weight_kg,
		   yft_n = yft_sp_nb,
		   yft_n_est = yft_sp_nb_est,
		   skj_w = skj_sp_weight_kg,
		   skj_n = skj_sp_nb,
		   skj_n_est = skj_sp_nb_est,
		   bet_w = bet_sp_weight_kg,
		   bet_n = bet_sp_nb,
		   bet_n_est = bet_sp_nb_est,
		   alb_w = alb_sp_weight_kg,
		   alb_n = alb_sp_nb,
		   alb_n_est = alb_sp_nb_est,
		   oth_w = oth_sp_weight_kg,
		   oth_n = oth_sp_nb,
		   oth_n_est = oth_sp_nb_est,
		   E = hooks_nb,
		   hbf = hooks_between_floats,
		   ymd = set_date)

# Convert units: weight to metric tons, effort to hundreds of hooks
EC_raw <- EC_raw |> 
	mutate(across(c(E, hbf, yft_n, yft_w, skj_n, skj_w, 
					bet_n, bet_w, alb_n, alb_w, 
					oth_n, oth_w), as.numeric),
		   yft_w = yft_w/1000, 
		   skj_w = skj_w/1000,
		   bet_w = bet_w/1000,
		   alb_w = alb_w/1000,
		   oth_w = oth_w/1000,
		   E = E/100,
		   ymd = as.Date(ymd, "%Y-%m-%d"))
```

```{r, eval = F}
# test with observers data
EC_raw <- read.csv("/home/romainf/Documents/FisheriesDatabases/operational_observer_logbook_catch_effort_data_yft_bet_skj_oth.csv")
stat_list$nrow_init <- nrow(EC_raw)

#name convention
EC_raw <- EC_raw |> 
	rename(flag = flag_id,
		   latitude = lat,
		   longitude = lon,
		   hbf = hook_between_floats,
		   yft_n = yft_nb,
		   yft_w = yft_kg_est,
		   skj_n = skj_nb,
		   skj_w = skj_kg_est,
		   bet_n = bet_nb,
		   bet_w = bet_kg_est,
		   alb_n = alb_nb,
		   alb_w = alb_kg_est,
		   oth_n = oth_nb,
		   oth_w = oth_kg_est,
		   E = hook_set,
		   ymd = set_date)

# Convert units: weight to metric tons, effort to hundreds of hooks
EC_raw <- EC_raw |> 
	mutate(across(c(E, hbf, yft_n, yft_w, skj_n, skj_w, 
					bet_n, bet_w, alb_n, alb_w, 
					oth_n, oth_w), as.numeric),
		   yft_w = yft_w/1000, 
		   skj_w = skj_w/1000,
		   bet_w = bet_w/1000,
		   alb_w = alb_w/1000,
		   oth_w = oth_w/1000,
		   E = E/100,
		   ymd = as.Date(ymd, "%Y-%m-%d"))
```

## Convert coordinates from degree-minutes to decimal degrees

```{r}

res <- convert_coordinates_df(EC_raw, report = TRUE)
EC_raw <- res$data
print(res$summary)
# View(res$failures)

EC_raw <- EC_raw |> 
	mutate(
		latCent = roundTo.5(latitude),
		lonCent = roundTo.5(longitude)
	)
```

## Remove entries with NA, NaN, Inf, or NULL in required fields

```{r}
required_cols <- c("ymd", "latitude", "longitude", "E", "yft_n", "skj_n", 
				   "bet_n", "alb_n", "oth_n")

EC_raw <- EC_raw |>
	filter(if_all(all_of(required_cols), is_valid))

stat_list$nrow_noNA <- nrow(EC_raw)
stat_list$outliers_NA <- stat_list$nrow_init - stat_list$nrow_noNA

report_removal(stat_list$outliers_NA, stat_list$nrow_init, "corrupted data")

cat("There are", stat_list$nrow_noNA, "entries available for", 
	sum(EC_raw$yft_n), "yellowfins caught\n")

timeframe <- range(EC_raw$ymd)
cat("Data ranges from", as.character(timeframe[1]), "to", 
	as.character(timeframe[2]), "\n\n")



```

# Handle Hooks Between Floats (HBF) quality issues


```{r}
# Track initial HBF statistics
hbf_threshold <- 50

stat_list$hbf_gt_threshold <- sum(EC_raw$hbf > hbf_threshold, na.rm = TRUE)
stat_list$hbf_le0 <- sum(EC_raw$hbf <= 0, na.rm = TRUE)

# Flag outliers > 50
EC_raw <- EC_raw |>
	mutate(hbf_outlier = hbf > hbf_threshold)

if(stat_list$hbf_gt_threshold > 0) {
	cat(stat_list$hbf_gt_threshold, " entries flagged with HBF > 50 (",
		round(stat_list$hbf_gt_threshold / nrow(EC_raw) * 100, 2), "% of data)\n")
}

# Convert invalid HBF (≤ 0) to NA
EC_raw <- EC_raw |>
	mutate(hbf = ifelse(hbf <= 0, NA, hbf))

if(stat_list$hbf_le0 > 0) {
	cat(stat_list$hbf_le0, " entries with HBF ≤ 0 converted to NA (",
		round(stat_list$hbf_le0 / nrow(EC_raw) * 100, 2), "% of data)\n")
}

# Summary
stat_list$hbf_valid <- sum(!is.na(EC_raw$hbf) & !EC_raw$hbf_outlier, na.rm = TRUE)
cat("HBF summary: ", stat_list$hbf_valid, " valid,",
	stat_list$hbf_le0, " NA,", 
	stat_list$hbf_gt_threshold, " outliers (>50)\n\n")
```

## Remove dummy fisheries (positive catch with zero effort)

```{r}
dummy_fisheries <- EC_raw |> filter(yft_n > 0 & E == 0)
stat_list$dummy_fisheries <- nrow(dummy_fisheries)

# if(stat_list$dummy_fisheries > 0) {
#   saveRDS(dummy_fisheries, "processed/EC_dummy_fisheries.RDS")
# }

report_removal(stat_list$dummy_fisheries, stat_list$nrow_init, 
			   "dummy fisheries (catch > 0, effort = 0)")

EC_raw <- EC_raw |> filter(!(yft_n > 0 & E == 0))
```

## Remove zero fisheries (zero catch AND zero effort)

```{r}
zero_fisheries <- EC_raw |> filter(yft_n == 0 & E == 0)
stat_list$zero_fisheries <- nrow(zero_fisheries)

# if(stat_list$zero_fisheries > 0) {
#   saveRDS(zero_fisheries, "processed/EC_zero_fisheries.RDS")
# }

report_removal(stat_list$zero_fisheries, stat_list$nrow_init, 
			   "zero fisheries (catch = 0, effort = 0)")

EC_raw <- EC_raw |> filter(!(yft_n == 0 & E == 0))
```

## Remove effort outliers (beyond 3 standard deviations)

```{r}
nrow_before <- nrow(EC_raw)

EC_raw <- EC_raw |>
	filter(abs(E - mean(E, na.rm = TRUE)) <= 3 * sd(E, na.rm = TRUE))

stat_list$outliers_effort <- nrow_before - nrow(EC_raw)
report_removal(stat_list$outliers_effort, stat_list$nrow_init, 
			   "effort outliers")
```

## Clean vessel names (remove trailing spaces)

```{r}
EC_raw <- EC_raw |>
	mutate(vesselname = str_trim(vesselname, side = "right"))
```

## Remove exact duplicate entries

```{r}
nrow_before <- nrow(EC_raw)
EC_raw <- EC_raw |> distinct()
stat_list$duplicates <- nrow_before - nrow(EC_raw)

report_removal(stat_list$duplicates, stat_list$nrow_init, "duplicates")
```

# Apply Pacific Ocean mask

```{r}
EC_masked <- EC_raw |>
	filter(latCent >= lat_range[1] & latCent <= lat_range[2], 
		   lonCent >= lon_range[1] & lonCent <= lon_range[2]) |>
	semi_join(pmask_lookup, by = c("latCent", "lonCent"))

mask_outliers <- anti_join(EC_raw, EC_masked, 
						   by = c("latCent", "lonCent", "ymd", "flag"))
stat_list$outliers_mask <- nrow(mask_outliers)

# if(stat_list$outliers_mask > 0) {
#   saveRDS(mask_outliers, "processed/EC_outliers_mask.RDS")
# }

report_removal(stat_list$outliers_mask, stat_list$nrow_init, 
			   "outside Pacific mask")

EC_raw <- EC_masked
stat_list$nrow_clean <- nrow(EC_raw)
```

```{r}
saveRDS(EC_raw, here("outputs/EC_data.RDS"))
```


## Data Cleaning Summary
```{r cleaning_summary, echo = F}
# Create summary table of all cleaning steps
cleaning_summary <- data.frame(
	Step = c(
		"Initial entries",
		"Corrupted data (NA/NaN/Inf)",
		paste0("HBF outliers (>", hbf_threshold, ")"),
		"Dummy fisheries (catch>0, effort=0)",
		"Zero fisheries (catch=0, effort=0)",
		"Effort outliers",
		"Exact duplicates",
		"Outside Pacific mask",
		"Final clean entries"
	),
	entries = c(
		stat_list$nrow_init,
		paste0(stat_list$outliers_NA),
		paste0(stat_list$hbf_gt_threshold),
		paste0(stat_list$dummy_fisheries),
		paste0(stat_list$zero_fisheries),
		paste0(stat_list$outliers_effort),
		paste0(stat_list$duplicates),
		paste0(stat_list$outliers_mask),
		stat_list$nrow_clean
	),
	Percent_of_Initial = c(
		100,
		round(stat_list$outliers_NA / stat_list$nrow_init * 100, 2),
		round(stat_list$hbf_gt_threshold / stat_list$nrow_init * 100, 2),
		round(stat_list$dummy_fisheries / stat_list$nrow_init * 100, 2),
		round(stat_list$zero_fisheries / stat_list$nrow_init * 100, 2),
		round(stat_list$outliers_effort / stat_list$nrow_init * 100, 2),
		round(stat_list$duplicates / stat_list$nrow_init * 100, 2),
		round(stat_list$outliers_mask / stat_list$nrow_init * 100, 2),
		round(stat_list$nrow_clean / stat_list$nrow_init * 100, 2)
	)
)

# Display table
knitr::kable(
	cleaning_summary,
	format = "simple",
	col.names = c("Cleaning Step", "Entries", "% of Initial"),
	caption = "Data Cleaning Summary Statistics",
	align = c("l", "r", "r")
)
```

### Key Statistics
```{r key_stats}
cat("Total data removed:", stat_list$nrow_init - stat_list$nrow_clean, "entries\n")
cat("Percentage removed:", 
	round((stat_list$nrow_init - stat_list$nrow_clean) / stat_list$nrow_init * 100, 2), "%\n\n")

cat("Final dataset:\n")
cat("  entries:", stat_list$nrow_clean, "\n")
# cat("  Yellowfin caught:", format(stat_list$catch_clean, big.mark = ","), "\n")
cat("  Date range:", as.character(min(EC_raw$ymd)), "to", 
	as.character(max(EC_raw$ymd)), "\n")
cat("  Spatial extent:", 
	round(min(EC_raw$latitude), 1), "to", round(max(EC_raw$latitude), 1), "°N,",
	round(min(EC_raw$longitude), 1), "to", round(max(EC_raw$longitude), 1), "°E\n")
```

TODO

check if it works with raised data


