---
title: 02 Processing operational data
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02 Processing Operational Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
```{r, warning=FALSE, message=F, echo=FALSE}
library(seafisheries)
library(dplyr)
library(stringr)
stat_list <- list()
pmask_lookup <- readRDS(system.file("extdata/outputs/pmask_lookup.RDS",
                                     package = "seafisheries"))
library(here)
dir.create(here("outputs"), showWarnings = FALSE, recursive = TRUE)

```
This document shows the process of formatting operational data to something usable with models. Using as example the yellowfin captain's and observer's logbooks

Summary:

- converts date/long/lat to usable format

- removes NA in the 3 dimensions and effort

- aggregates to 1x1

- crops to pacific region

- splits data between positive catch and effort / 0 catch and 0 effort / positive catch and 0 effort

# Simple example

This uses a fake data set of longline data. This especially contains hooks between floats so skip the hbf outlier section if using other type of data

## Load data and rename columns

```{r}
EC_raw <- readRDS(system.file("extdata/mockData/operational_data.RDS", 
                               package = "seafisheries"))
# EC_raw <- read.csv("/home/romainf/Documents/FisheriesDatabases/operational_captain_longline_logsheet_catch_effort_data_yft_bet_skj_oth.csv")
stat_list$nrow_init <- nrow(EC_raw)

# Rename columns to standard nomenclature
EC_raw <- EC_raw |> 
	rename(yft_w = yft_sp_weight_kg,
		   yft_n = yft_sp_nb,
		   yft_n_est = yft_sp_nb_est,
		   skj_w = skj_sp_weight_kg,
		   skj_n = skj_sp_nb,
		   skj_n_est = skj_sp_nb_est,
		   bet_w = bet_sp_weight_kg,
		   bet_n = bet_sp_nb,
		   bet_n_est = bet_sp_nb_est,
		   alb_w = alb_sp_weight_kg,
		   alb_n = alb_sp_nb,
		   alb_n_est = alb_sp_nb_est,
		   oth_w = oth_sp_weight_kg,
		   oth_n = oth_sp_nb,
		   oth_n_est = oth_sp_nb_est,
		   E = hooks_nb,
		   hbf = hooks_between_floats,
		   ymd = set_date)

# Convert units: weight to metric tons, effort to hundreds of hooks
EC_raw <- EC_raw |> 
	mutate(across(c(E, hbf, yft_n, yft_w, skj_n, skj_w, 
					bet_n, bet_w, alb_n, alb_w, 
					oth_n, oth_w), as.numeric),
		   yft_w = yft_w/1000, 
		   skj_w = skj_w/1000,
		   bet_w = bet_w/1000,
		   alb_w = alb_w/1000,
		   oth_w = oth_w/1000,
		   E = E/100,
		   ymd = as.Date(ymd, "%Y-%m-%d"))
```

```{r, eval = F}
# test with observers data
EC_raw <- read.csv("/home/romainf/Documents/FisheriesDatabases/operational_observer_logbook_catch_effort_data_yft_bet_skj_oth.csv")
stat_list$nrow_init <- nrow(EC_raw)

#name convention
EC_raw <- EC_raw |> 
	rename(flag = flag_id,
		   latitude = lat,
		   longitude = lon,
		   hbf = hook_between_floats,
		   yft_n = yft_nb,
		   yft_w = yft_kg_est,
		   skj_n = skj_nb,
		   skj_w = skj_kg_est,
		   bet_n = bet_nb,
		   bet_w = bet_kg_est,
		   alb_n = alb_nb,
		   alb_w = alb_kg_est,
		   oth_n = oth_nb,
		   oth_w = oth_kg_est,
		   E = hook_set,
		   ymd = set_date)

# Convert units: weight to metric tons, effort to hundreds of hooks
EC_raw <- EC_raw |> 
	mutate(across(c(E, hbf, yft_n, yft_w, skj_n, skj_w, 
					bet_n, bet_w, alb_n, alb_w, 
					oth_n, oth_w), as.numeric),
		   yft_w = yft_w/1000, 
		   skj_w = skj_w/1000,
		   bet_w = bet_w/1000,
		   alb_w = alb_w/1000,
		   oth_w = oth_w/1000,
		   E = E/100,
		   ymd = as.Date(ymd, "%Y-%m-%d"))
```

## Convert coordinates from degree-minutes to decimal degrees

```{r}

res <- convert_coordinates_df(EC_raw, report = TRUE)
EC_raw <- res$data
print(res$summary)
# View(res$failures)

EC_raw <- EC_raw |> 
	mutate(
		latCent = roundTo.5(latitude),
		lonCent = roundTo.5(longitude)
	)
```

## Remove invalid entries

`remove_invalid()` drops rows with NA, NaN, Inf, or NULL in any of the
columns you consider essential, here, coordinates, date, effort, and all
species catch counts.

```{r}
required_cols <- c("ymd", "latitude", "longitude", "E", "yft_n", "skj_n", 
				   "bet_n", "alb_n", "oth_n")

EC_raw <- remove_invalid(EC_raw, required_cols)
```

## Handle Hooks Between Floats (HBF); longline only

`treat_hbf()` is gear-specific. It flags HBF values above a threshold as
outliers (without removing rows) and converts non-positive HBF values to NA.
Skip this step entirely for gears without an HBF field.

```{r}
EC_raw <- treat_hbf(EC_raw, hbf_col = "hbf", threshold = 50)
```

## Recode school-association codes; purse seine only

`treat_school()` is the purse-seine equivalent of `treat_hbf()`. Valid school
types are 1-7; anything else present in the data (e.g. -9, -1, 0, 8) gets
recoded to 10 for later review, and missing values are left as NA. Shown for reference only.

```{r, eval = FALSE}
EC_raw <- treat_school(EC_raw, school_col = "school")
```

## Remove dummy fisheries

`remove_dummy_fisheries()` drops rows with positive catch (summed across
`catch_cols`) but zero effort. Set `return_dummy = TRUE` to get the removed
rows back alongside the cleaned data, e.g. for manual inspection.

```{r}
EC_raw <- remove_dummy_fisheries(EC_raw, 
								  catch_cols = c("yft_n", "skj_n", "bet_n", "alb_n", "oth_n"), 
								  effort_col = "E")
```

## Remove zero fisheries

`remove_zero_fisheries()` is the mirror case: rows with zero catch and zero
effort carry no information and are dropped.

```{r}
EC_raw <- remove_zero_fisheries(EC_raw, 
								 catch_cols = c("yft_n", "skj_n", "bet_n", "alb_n", "oth_n"), 
								 effort_col = "E")
```

## Remove effort outliers

`remove_outliers_sd()` drops rows more than `n_sd` standard deviations from
the mean of a numeric column -- here, effort.

```{r}
EC_raw <- remove_outliers_sd(EC_raw, col = "E", n_sd = 3)
```

## Clean vessel names

`trim_vesselname()` right-trims whitespace from a vessel name column. It
removes no rows, but has to run before duplicate removal, otherwise two
rows differing only by trailing whitespace on `vesselname` won't collapse
into a duplicate.

```{r}
EC_raw <- trim_vesselname(EC_raw, vesselname_col = "vesselname")
```

## Remove exact duplicates

`remove_duplicates()` drops exact duplicate rows across all columns.

```{r}
EC_raw <- remove_duplicates(EC_raw)
```

## Apply the Pacific Ocean mask

`apply_pacific_mask()` filters to the Pacific bounding box and semi-joins
against the mask lookup table, dropping grid cells outside the region of
interest.

```{r}
EC_raw <- apply_pacific_mask(EC_raw, lat_col = "latCent", lon_col = "lonCent", 
							  pmask_lookup = pmask_lookup)
```

## All of the above, in one call

`process_catch_data()` runs every step above in sequence (skipping
`treat_hbf`/`treat_school` unless `gear` is set, and skipping vessel-name
trimming unless `vesselname_col` is supplied). Equivalent to everything from
"Remove invalid entries" through "Apply the Pacific Ocean mask" above.
`return_dummy = TRUE` returns a list with `data` (cleaned) and
`dummy_fisheries` (the removed rows) instead of just the cleaned data.

```{r}
result <- process_catch_data(
	df              = EC_raw,
	required_cols   = required_cols,
	catch_cols      = c("yft_n", "skj_n", "bet_n", "alb_n", "oth_n"),
	effort_col      = "E",
	lat_col         = "latCent",
	lon_col         = "lonCent",
	pmask_lookup    = pmask_lookup,
	effort_sd_n     = 3,
	gear            = "L",
	hbf_col         = "hbf",
	hbf_threshold   = 50,
	vesselname_col  = "vesselname",
	return_dummy    = TRUE
)

EC_raw <- result$data
dummy_fisheries <- result$dummy_fisheries
```


```{r}
saveRDS(EC_raw, here("outputs/EC_data.RDS"))
```


## Data Cleaning Summary
```{r cleaning_summary, echo = F}
# Create summary table of all cleaning steps
cleaning_summary <- data.frame(
	Step = c(
		"Initial entries",
		"Corrupted data (NA/NaN/Inf)",
		paste0("HBF outliers (>", hbf_threshold, ")"),
		"Dummy fisheries (catch>0, effort=0)",
		"Zero fisheries (catch=0, effort=0)",
		"Effort outliers",
		"Exact duplicates",
		"Outside Pacific mask",
		"Final clean entries"
	),
	entries = c(
		stat_list$nrow_init,
		paste0(stat_list$outliers_NA),
		paste0(stat_list$hbf_gt_threshold),
		paste0(stat_list$dummy_fisheries),
		paste0(stat_list$zero_fisheries),
		paste0(stat_list$outliers_effort),
		paste0(stat_list$duplicates),
		paste0(stat_list$outliers_mask),
		stat_list$nrow_clean
	),
	Percent_of_Initial = c(
		100,
		round(stat_list$outliers_NA / stat_list$nrow_init * 100, 2),
		round(stat_list$hbf_gt_threshold / stat_list$nrow_init * 100, 2),
		round(stat_list$dummy_fisheries / stat_list$nrow_init * 100, 2),
		round(stat_list$zero_fisheries / stat_list$nrow_init * 100, 2),
		round(stat_list$outliers_effort / stat_list$nrow_init * 100, 2),
		round(stat_list$duplicates / stat_list$nrow_init * 100, 2),
		round(stat_list$outliers_mask / stat_list$nrow_init * 100, 2),
		round(stat_list$nrow_clean / stat_list$nrow_init * 100, 2)
	)
)

# Display table
knitr::kable(
	cleaning_summary,
	format = "simple",
	col.names = c("Cleaning Step", "Entries", "% of Initial"),
	caption = "Data Cleaning Summary Statistics",
	align = c("l", "r", "r")
)
```

### Key Statistics
```{r key_stats}
cat("Total data removed:", stat_list$nrow_init - stat_list$nrow_clean, "entries\n")
cat("Percentage removed:", 
	round((stat_list$nrow_init - stat_list$nrow_clean) / stat_list$nrow_init * 100, 2), "%\n\n")

cat("Final dataset:\n")
cat("  entries:", stat_list$nrow_clean, "\n")
# cat("  Yellowfin caught:", format(stat_list$catch_clean, big.mark = ","), "\n")
cat("  Date range:", as.character(min(EC_raw$ymd)), "to", 
	as.character(max(EC_raw$ymd)), "\n")
cat("  Spatial extent:", 
	round(min(EC_raw$latitude), 1), "to", round(max(EC_raw$latitude), 1), "°N,",
	round(min(EC_raw$longitude), 1), "to", round(max(EC_raw$longitude), 1), "°E\n")
```

TODO

check if it works with raised data


