---
title: 01 Setup
author: Romain Forestier
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01 Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F, warning = F, message = F, warning = FALSE}
source("../myLibrary.r")
knitr::opts_chunk$set(eval = T, echo = F, message = F, warning = F)

fig_width <- 20
fig_height <- 16

options(tibble.print_max = 50)

```


# Naming conventions


```{r, eval = F}
# not sure is needed at the moment
summary_table <- data.frame("name" = c("captain_LL_raw",
									   "captain_LL_1x1",
									   # "captain_LL_flags",
									   "captain_LL_outliers_mask",
									   "captain_LL_outliers_effort",
									   "captain_LL_ghostFisheries",
									   "captain_LL_beforeAgg",
									   "obs_LL_raw",
									   "obs_LL_1x1",
									   "obs_LL_flags",
									   "obs_LL_outliers_mask",
									   "obs_LL_outliers_effort",
									   "obs_LL_ghostFisheries",
									   "obs_LL_before_Agg",
									   "LL_1x1",
									   "aggregated_LL_5x5",
									   "delta_LL_5x5",
									   "aggregated_PS_1x1",
									   "CL_raw",
									   "CL_1x1",
									   "CL_correctLenght"),
							"content" = c("logsheets at .01 degrees resolution (hbf,hnb, nb, kg) for all species",
										  "logsheets aggregated to 1 degree resolution in the PO and summed over time (y-m)(effort, nb, kg)",
										  # "logsheets aggregated 1x1 with flag IDs",
										  "outliers not present in the pacific area",
										  "outliers with 0 catch and 0 effort",
										  "outliers with positive catch and 0 effort",
										  "logsheets with no outliers",
										  "logbooks at .01 degrees resolution (hbf,hnb, nb, kg)",
										  "logbooks aggregated to 1 degree resolution in the PO and summed over time (y-m-d)(effort, nb, kg)",
										  "logbooks aggregated 1x1 with flag IDs",
										  "logbooks not present in the pacific area",
										  "logbooks with effort anomalies",
										  "logbooks with catch without effort",
										  "logbooks with no outliers",
										  "aggregated database at 1 degrees of logbooks and logsheets",
										  "aggregated database at 5 degrees resolution without groups/sums",
										  "difference between LL_1x1 and aggregated database in 5x5 for LL",
										  "aggregated database at 1 degrees of PS",
										  "catch/length at .0001 degrees resolution in the PO",
										  "catch/length aggregated to 1 degree resolution with mean length",
										  "catch/length non aggregated with all length to UF")
							
)

knitr::kable(summary_table, caption = "Saved files", format = "html")
```


```{r}
# To redo when done with naming convention
keywords_table <- data.frame("name" = c("captain",
										"obs",
										"EC",
										"LF",
										"agg",
										"raw",
										"clean",
										"LL",
										"PS",
										"1x1",
										"5x5",
										"delta",
										"CL"),
							 "description" = c("captain logbooks",
							 				  "observer logbooks",
							 				  "Effort/Catch data",
							 				  "Length frequency data",
							 				  "aggregated",
							 				  "data format converted to usable values, no infomation removed (NULL,NA,rounding,...)",
							 				  "final result",
							 				  "longline fisheries",
							 				  "purse seine fisheries",
							 				  "data aggregated to the 1x1 resolution",
							 				  "data aggregated to the 5x5 resolution",
							 				  "difference between two databases",
							 				  "catch/length data")
							 
)



knitr::kable(keywords_table, caption = "Keywords")

```

```{r}
country_correspondence <- data.frame(
  code = c("AS"," AU", "BZ", "CK", "CN", "ES", "FJ", "FM", "FR", "ID", "JP", "KI", "KR", "MH", "NC", "NU", "NZ", "PF", "PG", "PH", "PT", "PW", "SB", "TO", "TV", "TW", "US", "VN", "VU", "WS"),
  country = c(
  	"American Samoa",
    "Australia",
    "Belize", 
    "Cook Islands",
    "China",
    "Spain",
    "Fiji",
    "Federated States of Micronesia",
    "France",
    "Indonesia", 
    "Japan",
    "Kiribati",
    "Korea (South)",
    "Marshall Islands",
    "New Caledonia",
    "Niue",
    "New Zealand", 
    "French Polynesia",
    "Papua New Guinea",
    "Philippines",
    "Portugal",
    "Palau",
    "Solomon Islands",
    "Tonga",
    "Tuvalu",
    "Taiwan",
    "United States",
    "Vietnam",
    "Vanuatu",
    "Samoa"
  )

)

knitr::kable(country_correspondence, caption = "Country flags")

```

# Naming comvention

yft = yellowfin

bet = bigeye tuna

alb = albacore

skj = skipjack

oth = other species than above tunas

n = number

w = weight in metric tons

set_date = date in yyyy-mm-dd ; to double check

set_date_mm = date in yyyy-mm-15

latitude/longitude = high resolution coordinate

latCent/lonCent = rounding coordinates to grid cell center

latSouth/lonWest = coordinates points at grid cell's South West corner (convention)

E = effort 


FLAG is the vessel nationality and FLEET is a subset within that FLAG – usually to distinguish what we refer to as (i) “distant-water” vessels that are typically larger, fish in the high seas for long periods and tranship at sea, and (ii) “offshore” “small-scale” vessels that fish for shorter periods and are based out their home country or in PIC ports.

# Masks 

## Land
The study area is determined by the mask below (ERA5 landmask). It corresponds to the climate forcings and model configurations. The corners of the Pacific model domain are 89E-71W, 55S-65N, the resolution 1 degree, so the mask dimension is 200x120.

```{r}
pmask <- scan("masks/mask_era5_np_1deg_po_cor_noIO.txt")
pmask <- matrix(pmask, ncol = 200, byrow = T)
rownames(pmask) <- lat_range[2]:lat_range[1]
colnames(pmask) <- lon_range[1]:lon_range[2]

plot_dat <- reshape2::melt(pmask)

p_mask <- ggplot(plot_dat) +
	geom_tile(aes(x = Var2, y = Var1, fill = as.factor(value)), width = 1, height = 1) +
	scale_fill_manual(values = c("chocolate3", "dodgerblue1","dodgerblue2","dodgerblue3")) +
	map_bg() +
	coord_labels +
	coord_fixed(xlim = lon_range, ylim = lat_range) +
	myTheme() +
	theme(legend.position = "none")

print(p_mask)
# ggsave("figures/map_mask.png", plot = p_mask, units = "cm", width = fig_width, height = fig_height)

# Pre-process pmask into a lookup dataframe for faster processing
pmask_lookup <- which(pmask > 0, arr.ind = TRUE) %>%
	as_tibble() %>%
	mutate(latCent = as.numeric(rownames(pmask)[row]),
		   lonCent = as.numeric(colnames(pmask)[col])) %>%
	dplyr::select(latCent, lonCent)
saveRDS(pmask_lookup,"outputs/pmask_lookup.RDS")

p5mask <- scan("masks/mask_era5_np_5deg_po_cor_noIO.txt")
p5mask <- matrix(p5mask, ncol = 40, byrow = T)
rownames(p5mask) <- seq(lat_range[2]-2, lat_range[1]-2, -5)
colnames(p5mask) <- seq(lon_range[1]-2, lon_range[2]-2, 5)
p5mask_lookup <- which(p5mask > 0, arr.ind = TRUE) %>%
	as_tibble() %>%
	mutate(latCent = as.numeric(rownames(p5mask)[row]),
		   lonCent = as.numeric(colnames(p5mask)[col])) %>%
	dplyr::select(latCent, lonCent)
```

## Temperature
An additional ERA5 temperature mask is applied to the data to remove all catch made in waters below 10 C. It works for the data up until the end of 2022.

```{r}
temp_dym <- dym::read.var.dym("masks/po_era5np_1x30d_T_L1_1959_2022.dym")
temp_arr <- temp_dym$var
dimnames(temp_arr) <- list("time" = as.character(temp_dym$t),
						   "longitude" = temp_dym$x,
						   "latitude" = temp_dym$y)
temp_mask <- reshape2::melt(temp_arr) %>% 
	# mutate(years = floor(time),
	# 	   months = floor((time - years) * 12) + 1,
	# 	   ymd = as.Date(paste0(years, "-", months, "-", 15))) %>% 
	# dplyr::select(-time,-years,-months) %>% 
	rename(temperature = value,
		   ymd = time)
```



